---
title: "Setup"
---

---

## `R` and RStudio

`R` is an open-source programming language for statistical computing.
`R` is widely-used in both academia and industry, due to its capacity for statistical analysis and data science.

In order to use `R` effectively, you need a suitable editing environment, i.e. RStudio.
For more context, RStudio is an integrated development environment (IDE) developed specifically for `R` programming.
Although `R` can be run without RStudio, RStudio provides a more user-friendly experience with additional functionality.

**To download `R` and RStudio, visit [`https://posit.co/download/rstudio-desktop`](https://posit.co/download/rstudio-desktop).**

You will see the following two tasks to be completed:

* **1: Install `R`**
* **2: Install RStudio**

## Step 1: Install `R`

(*Note that the following instructions apply to the latest `R` version (4.6.0) as of May 1, 2026*)

Click on ***DOWNLOAD AND INSTALL R***. This will direct you to the CRAN (Comprehensive R Archive Network) website.

**macOS**

* Click on ***Download R for macOS***.

* Choose the `.pkg` file suitable for your Mac (*Apple silicon (M1,2,..) Macs* or *older Intel Macs*.)

* Open the `.pkg` file after the download is complete.

* Follow the installation instructions.

**Windows**

* Click on ***Download R for Windows***.

* Click on ***install R for the first time*** (on the same line as the ***base*** subdirectory.)

* Choose ***Download R-4.6.0 for Windows***.

* Open the `.exe` file after the download is complete.

* Follow the installation instructions.

## Step 2: Install RStudio

::: callout-important
## Important

`R` **must** be installed before RStudio.
:::

Click on ***DOWNLOAD RSTUDIO DESKTOP...***
(If your operating system is not detected, scroll down and choose the right version.)

**macOS**

* Open the `.dmg` file after the download is complete.

* Drag and drop it to your Applications folder.

**Windows**

* Open the `.exe` file after the download is complete.

* Follow the installation instructions.

## Step 3: Check `R` and RStudio installations

Open RStudio and type in the following command in the **Console** pane.

```{r}
#| eval: false
version
```

This will print out the current version of `R` on your machine.
The output should look similar to what shown below (the first 4 lines might be different, depending on your operating system.)

```{r}
#| echo: false
version
```

## Step 4: Install an `R` package

In `R`, a package is a collection of functions, data, and compiled code.
In addition to a set of built-in base packages, there are numerous external `R` packages written by the community to add specific functionality.

In general, to install an `R` package, you can use the `install.packages()` function and pass in the package name.

The following example shows how to install the `tidyverse` package in `R`.
The `tidyverse` is a suite of `R` packages we will be using throughout this program.
It features popular packages such as `ggplot2` for data visualization and `dplyr` for data manipulation.

After installing `R` and RStudio, open RStudio and enter the following command in the **Console** pane.

```{r}
#| eval: false
install.packages("tidyverse")
```

To verify that `tidyverse` is successfully installed, run the following command:

```{r}
#| eval: false
library(tidyverse)
```

You should get a message similar to the output below.

```{r}
#| echo: false
library(tidyverse)
```

## Customizing RStudio

RStudio can be customized with different themes. To explore built-in themes,

* Navigate to the menu bar at the top of your screen

* Choose `Tools > Global Options... > Appearance`

* Change your RStudio theme under Editor theme

Note that within the Appearance tab, there are also options for changing your Editor font, Editor font size, etc.

Feel free to explore other options within the `Tools > Global Options...` menu.