---
title: "Homework 2"
subtitle: "36-315: Statistical Graphics and Visualization, Summer 2026"
author: "YOUR NAME HERE"
toc: true
fontsize: 10pt
geometry: margin=0.9in
format:
  pdf:
    colorlinks: true
execute:
  warning: false
  message: false
---

\newpage

# Problem 1: Statistical tests for categorical data (26 points)

For this problem, we will use the same Titanic dataset from Homework 1.

```{r}
library(tidyverse)
titanic <- read_csv("https://raw.githubusercontent.com/qntkhvn/36-315-summer26/refs/heads/master/data/titanic.csv")
```

## A (5pts) 

First, create a graph that shows the marginal distribution of the `Pclass` (cabin class) variable. Make sure that:

* Your axes are properly labeled

* The graph has a proper title

* Change the color of your plot so that it doesn't use default colors (automatically chosen by `R`).

```{r}
# YOUR CODE HERE
```

Next, describe the marginal distribution of `Pclass` in 1--2 sentences. In your answer, do not simply refer to "1", "2", and "3". Figure out what these categories mean using the data dictionary [here](https://www.kaggle.com/c/titanic/data).

**YOUR ANSWER HERE**

## B (5pts) 

Professor Ph.Dizzle looks at your graph in Part A and says: "The passengers were equally likely to belong to any of the three classes." You obviously disagree. You tell Professor Ph.Dizzle to just look at the graph, but they are still unconvinced (and appearing to get angry), arguing that what you see in the graph could just be due to random noise and may not be an actual significant difference.

From the graph alone, are you able to definitively state whether or not there is a statistically significant difference among the proportions of classes? State yes or no, and then provide a one-sentence explanation for your answer.

**YOUR ANSWER HERE**

## C (8pts) 

To provide more explicit evidence of statistical significance, what kind of statistical test could you use to show that the passengers are not equally likely to belong to any of the three classes?

**YOUR ANSWER HERE**

Now perform the statistical test, and formally interpret the results (in context). In particular:

* Display the code to run the test

* State the p-value from the test

* State your formal conclusion (in context) in 1--3 sentences.

```{r}
# YOUR CODE HERE
```

**YOUR ANSWER HERE**

## D (8pts) 

Now we wil make a couple of improvements to your graph in Part A. Using the code from Part A, make the following two changes:

* Annotate the p-value from Part C somewhere on your graph other than the title. If you got an extremely small p-value, it's fine if you describe the p-value as "Chi-squared test p-value approximately zero."

* Currently, your graph likely has the labels 1, 2, 3, which denote the levels for `Pclass`. Change these to "Upper", "Middle", "Lower", respectively, which are the proper class names.

```{r}
# YOUR CODE HERE
```

\newpage

# Problem 2: Visualizing uncertainty (49 points)

In this problem, we will work with a dataset of cereals manufactured in the United States. For a description of the variables, see [here](https://www.rdocumentation.org/packages/MASS/versions/7.3-65/topics/cereal).

```{r}
cereal <- read_csv("https://raw.githubusercontent.com/qntkhvn/36-315-summer26/refs/heads/master/data/cereal.csv")
```

## A (5pts) 

For this problem, we will focus on the manufacturer variable, `mfr`. First, produce a frequency table for the cereal manufacturers in the dataset. Then, answer the following: Which manufacturer is the most frequent, and which is the least frequent? In your answer, please include the names of the actual manufacturers, rather than just the initials.

```{r}
# YOUR CODE HERE
```

**YOUR ANSWER HERE**

## B (5pts) 

Create a frequency bar chart for `mfr`. Make the color of the bars something other than gray. Add appropriate labels and a title to your plot. Also, center the title by adding `+ theme(plot.title = element_text(hjust = 0.5))`. After making your plot, summarize it in 1--2 sentences.

```{r}
# YOUR CODE HERE
```

**YOUR ANSWER HERE**

## C (10pts) 

Now let's add confidence intervals to the plot in Part B. Write a function called `get_prop_ci()` that computes the 95% confidence intervals for the *proportion* of each manufacturer. Your function should take in a table object `freq_table` produced by the `table()` function.

```{r}
freq_table <- table(cereal$mfr)
```

To get you started, below is some template code. You need to fill in three lines of code for `get_prop_ci()` to do the following:

* Compute the sample size

* Compute the proportions of each category of `mfr`

* Compute the standard error of the proportions

```{r}
get_prop_ci <- function(freq_table, alpha = 0.05) {
  # compute sample size
  # n <-  
  # compute proportions
  # prop <- 
  # compute standard error for proportions
  # se <-  
  # compute confidence intervals
  # ci_lower <- prop - qnorm(1 - alpha / 2) * se
  # ci_upper <- prop + qnorm(1 - alpha / 2) * se
  # ci <- rbind(ci_lower, ci_upper)
  # return(ci)
}
```

After you've written `get_prop_ci()`, uncomment the following line of code to display confidence intervals.

```{r}
# get_prop_ci(freq_table)
```

Is there anything non-intuitive with any of the confidence intervals above? State Yes or No, and give a 1--2 sentence explanation.

**YOUR ANSWER HERE**

## D (3pts) 

Based on Part C, write one line of code that displays the 95% confidence intervals for the *counts* of each `mfr`.

(**Hint**: Remember that the proportion for any category $j$ is defined as $p_j = n_j / n$, where $n_j$ is the number of category $j$ and $n$ is the total sample size. You already have the confidence interval for $p_j$ and just need the confidence interval for $n_j$.)

```{r}
# YOUR CODE HERE
```

## E (10pts) 

Create a bar chart with the 95% confidence intervals.

Note that a bar chart with 95% confidence intervals ultimately displays only four pieces of information:

* The name of each category

* The counts of each category

* The lower bound of the 95% confidence interval for each category

* The upper bound of the 95% confidence interval for each category

Thus, first, we're going to make a new `data.frame` that appropriately summarizes the data we need for our plot:

```{r}
# store your answer to part D
# count_ci <- ?

# cereal_ci <- data.frame(mfr = ?, # name of each manufacturer
#                         counts = ?, # counts for each manufacturer
#                         lower = ?, # lower bound of 95% CI
#                         upper = ?) # upper bound of 95% CI
```

Your main goal for this part is to appropriately fill in the above "?". Don't just hard code the names/counts of each manufacturer. Figure out a way to use the code you've already written to define these variables. Because there are six manufacturers in the data, `cereal_ci` should only have six rows (one for each manufacturer).

(**Hint**: Part A should give you everything you need for `mfr` and `counts`, and Part D should give you everything you need for `lower` and `upper`.)

Once you've defined `cereal_ci` correctly, the following code should produce the desired bar chart. Uncomment the code, and also add color, appropriate labels, and a title to the bar chart.

```{r}
# cereal_ci |>
#   ggplot(aes(x = counts, y = mfr)) +
#   geom_col() +
#   geom_errorbar(aes(xmin = lower, xmax = upper))
```

## F (6pts) 

Now we're going to interpret the confidence intervals you visualized in Part E. Let's say that Professor Aretrash asks the following questions:

* Does it appear that there are any significant differences in counts among manufacturers that *are not* General Mills or Kellogg's?

**YOUR ANSWER HERE**

* For which manufacturers does General Mills appear to have significantly different counts?

**YOUR ANSWER HERE**

For the above questions, ignore issues of multiple testing; we will address that in the next part. And for each question, explain your reasoning in 1--3 sentences. **Hint**: When answering these questions, it may be helpful to look at the actual confidence intervals you computed in Part D.

## G (10pts) 

In Part F, Professor Aretrash asked you to make several pairwise comparisons between one manufacturer and another. You can find that there are 15 different pairwise comparisons you could make from your bar chart. As a result, if we attempt to make multiple pairwise comparisons from a single graph with 95% confidence intervals, our chance of making a Type 1 error is greater than 5%.

To address this issue: 

* **Make a bar chart similar to Part E, but with confidence intervals that incorporate the Bonferroni correction for 15 comparisons**. In other words, your graph here should be exactly the same as the graph in Part E, but with wider confidence intervals that incorporate the Bonferroni correction. 

* **Then, outside of your code, discuss (in 1-2 sentences) how you constructed the Bonferroni confidence intervals. After you've done that, answer the same two questions from Professor Aretrash's in Part F**, but using the graph you've made here.

**Hint**: In general, $\alpha$-level confidence intervals are constructed using the Normal quantile $z_{1-\alpha/2}$. When we construct 95\% confidence intervals, $\alpha = 0.05$, and thus we use $z_{1 - \alpha/2} = z_{0.975}$ (computed with `qnorm(0.975)`). This is exactly what we did with `get_prop_ci()` in Part C, where we set `alpha = 0.05` by default. When we use the Bonferroni correction, $\alpha$ is instead set to $0.05/k$, where $k$ is equal to the number of comparisons. For this part you need appropriately use `get_prop_ci()`, which you've already written, and then make a new graph.

```{r}
# YOUR CODE HERE
```

**YOUR ANSWER HERE**

**YOUR ANSWER HERE**

\newpage

# Problem 3: Resisting the first order for categories (25 points)

For this problem, we'll work with the [`starwars`](https://dplyr.tidyverse.org/reference/starwars.html) dataset, which is included in the `dplyr` package.

The code below loads the data, then uses the [`unnest`](https://tidyr.tidyverse.org/reference/nest.html) function to make a new dataset `starwars_character_films` where each row corresponds to a character appearance in a film. Type `help(starwars)` to view more information about the `starwars` dataset.

```{r}
starwars_character_films <- starwars |>
  dplyr::select(name, films) |>
  unnest(films)
```

We will focus on the `films` variable, which indicates the film a particular character appeared in. Specifically, we will demonstrate how to reorder categories on graphs in `R`. This is particularly useful when visualizing categorical variables, because the plotting order that `R` chooses by default may not be the best choice for your graphs.

## A (5pts)

First, make a bar chart that displays the frequency on the x-axis and `films` on the y-axis. Make sure your plot is properly titled and labeled, and choose a non-default color.

What is the default plotting order for categorical ("character" or "factor") variables in `R`?  (You should be able to determine this from glancing at the order of the categories in your plot, which starts from the bottom.)

```{r}
# YOUR CODE HERE
```

**YOUR ANSWER HERE**

## B (5pts)

Read this [introduction](https://forcats.tidyverse.org/) to the `forcats` package, which was designed to work with categorical data in `R`. The cheatsheet on that page will be particularly useful.

* Which function in the `forcats` package can be used to reorder the levels of a factor in any order you want?

**YOUR ANSWER HERE**

* Which function in the `forcats` package can be used to reorder the categories from most frequent to least frequent?

**YOUR ANSWER HERE**

* Which combination of functions in the `forcats` package can be used to reorder the categories from least frequent to most frequent?

**YOUR ANSWER HERE**

## C (5pts) 

Recreate the plot in Part A, but order the films either from least to most frequent or from most to least frequent.

```{r}
# YOUR CODE HERE
```

## D (5pts)

Recreate the plot in Part A, but order the films based on episode order (The Phantom Menace, Attack of the Clones, ..., The Force Awakens). See the Star Wars [wikipedia page](https://en.wikipedia.org/wiki/Star_Wars#The_Skywalker_saga) for the episode and release orders. 

```{r}
# YOUR CODE HERE
```

## E (5pts) 

Recreate the plot from part D, but rename the categories so that they use the following film abbreviations: I, II, III, IV, V, VI, VII. Again, see the Star Wars [wikipedia page](https://en.wikipedia.org/wiki/Star_Wars#The_Skywalker_saga) for more information.

```{r}
# YOUR CODE HERE
```