---
title: "Nonlinear regression and more quantitative data visualization"
subtitle: "<br>36-315: Statistical Graphics and Visualization, Summer 2026"
footer:  "[36-315-summer26.netlify.app](https://36-315-summer26.netlify.app)"
format:
  revealjs:
    embed-resources: true
    height: 900
    width: 1600
    theme: theme.scss
    smaller: true
    slide-number: c/t
    linestretch: 1.4
    code-line-numbers: false
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

## Beyond linear regression

* The truth is (almost) never linear

* But often the linearity assumption is good enough

* Many nonparametric/nonlinear regression methods

  * splines
  
  * **local regression**
  
  * generalized additive models
  
  * kernel regression
  
  * wavelets

## Nonlinear regression

* Assume normality, but not linearity where $f(x)$ is some unknown function $$Y_i \stackrel{iid}{\sim} N(\underbrace{f(X_i)}, \sigma^2)$$

* Do not make explicit assumptions about the functional form of $f$ (i.e., $f$ is estimated from the data)

* **Intuition: Any nonlinear function is locally linear**

* Local linear regressions fits a bunch of... local linear regressions, and then glues them together

* Local linear regression is basically **weighted linear regression**, where only "local units" get weight

## Weighted linear regression

* In linear regression, we solve

$$\arg \min_{\beta_0,\beta_1} \sum_{i=1}^n (Y_i - \beta_0 - \beta_1 X_i)^2$$

. . .

* In **weighted** linear regression, we solve

$$\arg \min_{\beta_0,\beta_1} \sum_{i=1}^n w_i \cdot (Y_i - \beta_0 - \beta_1 X_i)^2$$

. . .

* Local linear regression is exactly the same, except the weights depend on which $x$ we want to estimate $f(x)$.

## Local linear regression via LOESS

In **local linear regression**, we estimate $f(X_i)$ and solve

$$\text{arg }\underset{\beta_0, \beta_1}{\text{min}} \sum_i^n w_i(x) \cdot \big(Y_i - \beta_0 - \beta_1 X_i \big)^2$$ 

* Notice the weights depend on $x$, so observations close to $x$ gets more weight in estimating $f(x)$

. . .

`geom_smooth()` uses tri-cubic weighting:

$$w_i(d_i) = \begin{cases} (1 - |d_i|^3)^3, \text{ if } i \in \text{neighborhood of  } x\\
0 \text{ if } i \notin \text{neighborhood of  } x \end{cases}$$

* $d_i$: distance between $x$ and $X_i$ scaled to be between 0 and 1
  
* `span`: controls fraction of observations in neighborhood around the observation of interest

  * Default: `span = 0.75` (use 75% of the observations to fit local linear regression with weights)
  
:::aside
LOESS (Locally Estimated Scatterplot Smoothing) or LOWESS (Locally Weighted Scatterplot Smoothing)
:::
  
## Local linear regression, animated

```{r}
#| echo: false
#| fig-width: 8
library(tidyverse)
library(gganimate)
library(broom)
set.seed(1234)
x <- seq(0, 10, 10/99)
y <- 0.7 * sin(0.9 * x) + 0.08 * (x - 5)^2 - 0.7 + rnorm(100, 0.7)
df <- tibble(x, y)
span <- 0.25
d <- df |>
  crossing(center = df$x) |>
  mutate(dist = abs(x - center)) |>
  filter(rank(dist) / n() <= span) |> 
  mutate(weight = (1 - (dist / max(dist)) ^ 3) ^ 3)
fit <- loess(y ~ x, degree = 1, span = span, data = df)
a <- fit |> 
  augment() |> 
  crossing(center = df$x) |> 
  filter(x <= center)
d |> 
  ggplot(aes(x, y)) +
  geom_point(data = df, alpha = 0.3, color = "gray70") +
  geom_point(aes(alpha = weight)) +
  geom_vline(aes(xintercept = center), linetype = "dashed", color = "grey40") +
  geom_smooth(aes(group = center, weight = weight), method = "lm", se = FALSE, color = "blue") +
  geom_line(data = a, aes(x, .fitted), color = "red", linewidth = 1) +
  scale_alpha_continuous(range = c(0, 1), guide = "none") +
  transition_states(center, transition_length = 1, state_length = 0) +
  theme_light(base_size = 14)
```


## Displaying trend lines: LOESS

::: columns
::: {.column}

```{r}
library(tidyverse)
theme_set(theme_light())
```

```{r}
#| echo: false
theme_set(theme_light(base_size = 20))
```

```{r}
#| eval: false
penguins |> 
  ggplot(aes(x = flipper_len, y = body_mass)) +
  geom_point(alpha = 0.5) + 
  geom_smooth()
```

* LOESS is default behavior when $n \leq 1000$

* Default `span = 0.75`

* For $n > 1000$, `mgcv::gam()` is used with<br>`formula = y ~ s(x, bs = "cs")` and<br>`method = "REML"`

:::
::: {.column}

```{r}
#| echo: false
#| fig-height: 9
penguins |> 
  ggplot(aes(x = flipper_len, y = body_mass)) +
  geom_point(size = 3, alpha = 0.5) + 
  geom_smooth()
```

:::
:::

## Adjusting `span`

```{r}
#| echo: false
#| fig-height: 7
#| fig-width: 9
set.seed(1234)
x <- runif(100, 0, 10)
y <- sin(x) + rnorm(100, 0, 0.3)

library(gganimate)
tibble(span = seq(0.05, 1, by = 0.05)) |>
  group_by(span) |>
  do(broom::augment(loess(y ~ x, degree = 2, span = .$span))) |> 
  ggplot(aes(x, y)) +
  geom_point(color = "blue", size = 2) +
  geom_line(aes(y = .fitted), color = "chocolate", linewidth = 2) +
  transition_states(span, transition_length = 1, state_length = 0) +
  ggtitle("span = {closest_state}") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Pairs plots

* Create a pairs plot to see all pairwise relationships in one plot

* Pairs plot can include the various kinds of pairwise plots

  * Two quantitative variables: scatterplot

  * One categorical, one quantitative: side-by-side violins, stacked histograms, overlaid densities

  * Two categorical: stacked bars, side-by-side bars, mosaic plots
  
## Pairs plots 

```{r}
#| fig-height: 9
#| fig-width: 15
library(GGally)
penguins |> 
  select(bill_len, bill_dep, flipper_len, body_mass) |> 
  ggpairs()
```

## Pair plots of quantitative variables

::: columns
::: {.column width="42%"}

* Diagonal: marginal distributions

* Off-diagonal: joint (pairwise) distributions or statistical summaries (e.g., correlation)

* Matrix of plots is symmetric

```{r}
#| eval: false
penguins |> 
  select(bill_len, bill_dep, flipper_len, body_mass) |> 
  ggpairs(aes(alpha = 0.2))
```

Syntax note: change `alpha` within `aes` (annoying...)

:::
::: {.column width="58%"}

```{r}
#| echo: false
#| fig-height: 9
penguins |> 
  select(bill_len, bill_dep, flipper_len, body_mass) |> 
  ggpairs(aes(alpha = 0.2))
```

:::
:::

## Pair plots with categorical variables

::: columns
::: {.column width="43%"}

* Can either use `select()` before plotting or specify `columns` directly

```{r}
#| eval: false
penguins |> 
  ggpairs(columns = c("bill_len", "body_mass", "island"),
          aes(alpha = 0.2, color = species))
```

:::
::: {.column width="57%"}

```{r}
#| echo: false
#| fig-height: 9
penguins |> 
  ggpairs(columns = c("bill_len", "body_mass", "island"),
          aes(alpha = 0.2, color = species))
```

:::
:::

## More customization

::: columns
::: {.column width="38%"}

* Can customize which figures are displayed in which panels

* For more details, check out the package documentation [here](https://ggobi.github.io/ggally/articles/ggpairs.html)

```{r}
#| eval: false
penguins |> 
  ggpairs(
    columns = c("bill_len", "body_mass", "island"),
    mapping = aes(alpha = 0.2, color = species), 
    lower = list(continuous = "smooth_lm", 
                 combo = "facetdensitystrip"),
    upper = list(continuous = "cor", 
                 combo = "facethist")
  )
```

:::
::: {.column width="62%"}

```{r}
#| echo: false
#| fig-height: 9
penguins |> 
  ggpairs(
    columns = c("bill_len", "body_mass", "island"),
    mapping = aes(alpha = 0.2, color = species), 
    lower = list(continuous = "smooth_lm", 
                 combo = "facetdensitystrip"),
    upper = list(continuous = "cor", 
                 combo = "facethist")
  ) +
  theme_light(base_size = 16)
```

:::
:::

## More customization

::: columns
::: {.column width="46%"}

* Can also incorporate `ggplot2` customization

```{r}
#| eval: false
penguins |>
  ggpairs(columns = c("bill_len", "bill_dep", "flipper_len"),
          aes(color = species, alpha = 0.2)) +
  scale_color_manual(values = c("orange", "blue", "cyan4")) +
  scale_fill_manual(values = c("orange", "blue", "cyan4")) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

:::
::: {.column width="54%"}

```{r}
#| echo: false
#| fig-height: 8
penguins |>
  ggpairs(columns = c("bill_len", "bill_dep", "flipper_len"),
          aes(color = species, alpha = 0.2)) +
  scale_color_manual(values = c("orange", "blue", "cyan4")) +
  scale_fill_manual(values = c("orange", "blue", "cyan4")) +
  theme_minimal(base_size = 16) +
  theme(strip.text = element_text(face = "bold"))
```

:::
:::

## Correlation heatmaps

::: columns
::: {.column}

* Use `correlate()` and `scretch()` functions from the `corrr` package to compute correlation matrix and pivot

* Use `geom_tile()` to create a heatmap (tile map)

```{r}
library(corrr)
```

```{r}
#| eval: false
penguins |> 
  select(bill_len, bill_dep, flipper_len, body_mass) |> 
  correlate(diagonal = 1) |> 
  stretch() |> 
  ggplot(aes(x, y, fill = r)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(high = "red", mid = "white", low = "blue", 
                       midpoint = 0, limits = c(-1, 1)) +
  coord_fixed()
```

* See also: [`ggcorrplot`](https://rpkgs.datanovia.com/ggcorrplot)

:::
::: {.column}

<br>

```{r}
#| echo: false
#| fig-height: 7
penguins |> 
  select(bill_len, bill_dep, flipper_len, body_mass) |> 
  correlate(diagonal = 1) |> 
  stretch() |> 
  ggplot(aes(x, y, fill = r)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(high = "red", mid = "white", low = "blue", 
                       midpoint = 0, limits = c(-1, 1)) +
  coord_fixed()
```


:::
:::

## Parallel coordinates plots

::: columns
::: {.column width="42%"}

* Display each variable side-by-side on standardized axis

* Connect observations with lines

* Useful when working with a moderate number of observations and variables<br>(can be overwhelming with too many)

* Adjust the transparency of the lines to handle overlap

```{r}
#| eval: false
library(GGally)
penguins |>
  select(bill_len, bill_dep, flipper_len, body_mass) |> 
  ggparcoord(alphaLines = 0.2)
```

:::
::: {.column width="58%"}

```{r}
#| echo: false
#| fig-height: 8
library(GGally)
penguins |>
  select(bill_len, bill_dep, flipper_len, body_mass) |> 
  ggparcoord(alphaLines = 0.2)
```

:::
:::

## Parallel coordinates plots

::: columns
::: {.column width="45%"}

* Color lines by a categorical variable

* Can be useful for revealing group structure

* How do we order the x-axis?

```{r}
#| eval: false
penguins |>
  ggparcoord(columns = 3:6, alphaLines = 0.2, 
             groupColumn = "species", order = c(6, 5, 3, 4)) +
  theme(legend.position = "top")
```

:::
::: {.column width="55%"}

```{r}
#| echo: false
#| fig-height: 9
penguins |>
  ggparcoord(columns = 3:6, alphaLines = 0.2, 
             groupColumn = "species", order = c(6, 5, 3, 4)) +
  theme(legend.position = "top")
```

:::
:::

## Parallel coordinates plots

::: columns
::: {.column width="45%"}

* By default, the y-axis is scaled by subtracting the mean and dividing by the standard deviation

* Use `scale = "uniminmax"` so that minimum of the variable is 0 and the maximum is 1

```{r}
#| eval: false
penguins |>
  ggparcoord(columns = 3:6, alphaLines = 0.2, 
             groupColumn = "species", scale = "uniminmax") +
  theme(legend.position = "top")
```

:::
::: {.column width="55%"}

```{r}
#| echo: false
#| fig-height: 9
penguins |>
  ggparcoord(columns = 3:6, alphaLines = 0.2, 
             groupColumn = "species", scale = "uniminmax") +
  theme(legend.position = "top")
```

:::
:::