---
title: "Principal component analysis"
subtitle: "<br>36-315: Statistical Graphics and Visualization, Summer 2026"
footer:  "[36-315-summer26.netlify.app](https://36-315-summer26.netlify.app)"
format:
  revealjs:
    embed-resources: true
    height: 900
    width: 1600
    theme: theme.scss
    smaller: true
    slide-number: c/t
    linestretch: 1.4
    code-line-numbers: false
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

## Dimension reduction: the big picture

* **Key question: How do we visualize the structure of high-dimensional data?**

. . .

* Example: What if you're given a dataset with 50 variables and are asked to make one visualization that best represents the data? What do you do?

. . .

* Tedious task: Make a series of all $\displaystyle \binom{50}{2} = 1225$ pairs of plots? Or make a giant correlation heatmap?

. . .

* **Intuition: Take high-dimensional data and represent it in 2--3 dimensions, then visualize those dimensions**


## Dimension reduction: the big picture

* It's usually really hard to visualize many dimensions at the same time

. . .

* Often, it makes a lot of sense to choose 2--3 of the "most important dimensions" and just plot those

. . .

* PCA is a very common way to define "most important dimensions"

. . .

* PCA provides the linear combinations of variables that capture the most variation in the data

. . .

* It's common to plot the first two principal components in a scatterplot

. . .

* It's very useful to plot principal components with a biplot

  * Adds interpretability to the principal components, and helps reveal relationships among the variables

## What is the goal of dimension reduction?

Given a dataset with $p$ variables (columns) for $n$ observations (rows), can we find a smaller number of dimensions that captures interesting structure in the data?

. . .

**Dimension reduction**: 

* Focus on reducing the dimensionality of the feature space (i.e., number of columns) 

* **Retain** most of the information/**variability** in a lower dimensional space (i.e., reducing the number of columns)

## Principal components analysis (PCA)

* PCA produces a low-dimensional representation of a dataset

* PCA finds a set of variables that are linear combinations of the features that have maximal variance *and* are mutually uncorrelated

* The set of uncorrelated variables is called the **principal components (PCs)**

## Principal components

* PCs are weighted, normalized linear combinations of the original variables

* Weights reveal how different variables are **loaded** into the PCs

* First PC accounts for most variation in the data, second PC for second-most variation, and so on

## Principal components (a little more formal)

* Assume $\boldsymbol{X}$ is a $n \times p$ matrix that is **centered** and **stardardized**

. . .

* PCA will give us $p$ principal components that are $n$-length columns, denoted by $Z_1, \dots, Z_p$

  * The first PC is the linear combination that has the largest possible variance

  * Each succeeding component has the largest possible variance under the constraint that it is uncorrelated with the preceding components

  * **A small number of principal components often explains a high proportion of the variability in the data**

## First principal component

$$Z_1 = \phi_{11} X_1 + \phi_{21} X_2 + \dots + \phi_{p1} X_p$$

. . .

* $\phi_{11}, \phi_{21}, \dots, \phi_{p1}$ are the loadings of the first PC 

* These are weights indicating the contributions of each feature
  
* Weights are normalized $\displaystyle \sum_{j=1}^p \phi_{j1}^2 = 1$ (else, arbitrarily large variance)
  
* Together, the loadings make up the **loading vector** for the first PC, $\phi_{1} = (\phi_{11}, \phi_{21}, \dots, \phi_{p1})$

* $Z_1$ is a linear combination of the $p$ variables that has the **largest variance** (among all PCs)

## Second principal component

$$Z_2 = \phi_{12} X_1 + \phi_{22} X_2 + \dots + \phi_{p2} X_p$$

* $\phi_{12}, \phi_{22}, \dots, \phi_{p2}$ are the loadings of the second PC 
  
* Weights are normalized $\displaystyle \sum_{j=1}^p \phi_{j2}^2 = 1$
  
* Together, the loadings make up the **loading vector** for the second PC, $\phi_{2} = (\phi_{12}, \phi_{22}, \dots, \phi_{p2})$

* $Z_2$ is a linear combination of the $p$ variables that has the **largest variance** *but* **subject to constraint it is uncorrelated with $Z_1$**
    
## Obtaining the principal components

Repeat the same process to create $p$ principal components

* **Uncorrelated**: Each ($Z_j, Z_{j'}$) is uncorrelated with each other

* **Ordered Variance**: $\text{Var}(Z_1) > \text{Var}(Z_2) > \dots > \text{Var}(Z_p)$

* **Total Variance**: $\displaystyle \sum_{j=1}^p \text{Var}(Z_j) = p$

PCA can be

* formulated as an optimization problem

* solved via linear algebra techniques (e.g. eigen decomposition or singular value decomposition (SVD))

## Geometry of PCA

* The loading vector $\phi_1$ with elements $\phi_{11}, \phi_{21}, \dots, \phi_{p1}$ defines a direction in feature space along which the data vary the most

* If we project the observations in our dataset onto this direction, the projected values are called the principal component scores

<center>

<img src= "/images/pca-2d.png" style="border: 1px solid grey;" height="480">

</center>

## Geometry of PCA in 2D

```{r}
#| echo: false
#| fig-height: 7
library(tidyverse)
theme_set(theme_light(base_size = 20))
set.seed(123) 
plot_data <- tibble(x = rnorm(50, mean = 100, sd = 20)) |>
  mutate(y =  0.8 * x + rnorm(50, mean = 0, sd = 10))
basic_scatter <- ggplot(plot_data) +
  geom_point(aes(x, y), color = "black")+
  coord_equal()
basic_scatter
```

## Geometry of PCA in 2D

```{r}
#| echo: false
#| fig-height: 7
line1 <- lm(y ~ x, plot_data)$coef
line1_slope <- line1[2]
line1_intercept <- line1[1]
basic_scatter_yfit <- basic_scatter +
  geom_abline(aes(slope = line1_slope, intercept = line1_intercept),
              colour = "darkorange", linewidth = 1.5) +
  annotate("text", x = 75, y = 115, size = 11,
           label = "y ~ x", color = "darkorange")
basic_scatter_yfit
```

## Geometry of PCA in 2D

```{r}
#| echo: false
#| fig-height: 7
line2 <- lm(x ~ y, plot_data)$coef
line2_slope <- 1 / line2[2]
line2_intercept <- -(line2[1] / line2[2])
basic_scatter_xyfit <- basic_scatter_yfit +
  geom_abline(aes(slope = line2_slope, intercept = line2_intercept),
              colour = "blue", linewidth = 1.5) +
  annotate("text", x = 125, y = 55, size = 11,
           label = "x ~ y", color = "blue")
basic_scatter_xyfit
```

## Geometry of PCA in 2D

```{r}
#| echo: false
#| fig-height: 7
pca <- prcomp(cbind(plot_data$x, plot_data$y))$rotation
pca_slope <- pca[2,1] / pca[1,1]
pca_intercept <- mean(plot_data$y) - (pca_slope * mean(plot_data$x))

basic_scatter_y_pca_fit <- basic_scatter_xyfit +
  geom_abline(aes(slope = pca_slope, intercept = pca_intercept),
              colour = "darkgreen", linewidth = 1.5) +
  annotate("text", x = 135, y = 95, size = 11,
           label = "PC1", color = "darkgreen")
basic_scatter_y_pca_fit
```

## Visualizing PCA in two dimensions

Key idea: provide low-dimensional linear surfaces that are closest to the observations

```{r}
#| echo: false
#| fig-width: 15
#| fig-height: 6
plot_data |>
  mutate(
    yhat_line1 = (x * line1_slope + line1_intercept),
    xhat_line1 = x,
    yhat_line2 = y,
    xhat_line2 = (y - line2_intercept) / line2_slope,
    a = pca_slope,
    b = -1,
    c = pca_intercept,
    xhat_line3 = (b * (b * x - a * y) - (a * c)) / ((a * a) + (b * b)),
    yhat_line3 = (a * (-b * x + a * y) - (b * c)) / ((a * a) + (b * b)),
    slope_line1 = line1_slope,
    slope_line2 = line2_slope,
    slope_line3 = pca_slope,
    intercept_line1 = line1_intercept,
    intercept_line2 = line2_intercept,
    intercept_line3 = pca_intercept
  ) |> 
  select(-c(a, b, c)) |> 
  pivot_longer(yhat_line1:intercept_line3, names_to = "key", values_to = "value") |>
  separate(key, c("type", "line"), "_") |> 
  pivot_wider(names_from = "type", values_from = "value") |> 
  mutate(
    line = case_when(
      line=="line1" ~ "y ~ x",
      line=="line2" ~ "x ~ y",
      TRUE ~ "PC1"),
    line = fct_relevel(line, "y ~ x", "x ~ y", "PC1")
  ) |> 
  ggplot() +
  geom_point(aes(x = x, y = y, color = line)) +
  geom_abline(aes(slope = slope, intercept = intercept, color = line)) +
  geom_segment(aes(x = x, y = y, xend = xhat, yend = yhat, color = line)) +
  facet_wrap(~ line, ncol = 3) +
  scale_color_manual(values = c("darkorange", "blue", "darkgreen")) +
  theme(legend.position = "none",
        strip.text = element_text(size = 20, color = "black"),
        strip.background = element_rect(fill = NA))
```

. . .

* The above is the **minimizing projection residuals** viewpoint of PCA

* There's another viewpoint: **maximizing variance**. That is, if we project all points onto the solid green line, we maximize the variance of the resulting projected points across all such green lines

## PCA example: Starbucks drinks

```{r}
library(tidyverse)
theme_set(theme_light())
starbucks <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-12-21/starbucks.csv"
) |>
  # convert columns to numeric that were saved as character
  mutate(trans_fat_g = as.numeric(trans_fat_g), fiber_g = as.numeric(fiber_g))
glimpse(starbucks)
```

```{r}
#| echo: false
theme_set(theme_light(base_size = 20))
```

## Implement PCA

* Use `prcomp()` (based on SVD) for PCA on centered and scaled data

```{r}
starbucks_feat <- starbucks |> 
  select(serv_size_m_l:caffeine_mg)
starbucks_pca <- prcomp(starbucks_feat, center = TRUE, scale. = TRUE)
summary(starbucks_pca)
```

## Compute principal components

* Extract the matrix of principal components (dimension will match original data)

* Note: columns are uncorrelated, such that $\text{Var}(Z_1) > \text{Var}(Z_2) > \dots > \text{Var}(Z_p)$ 

```{r}
starbucks_pc_matrix <- starbucks_pca$x
head(starbucks_pc_matrix)
```

## Extract loading matrix

```{r}
starbucks_pca$rotation
```

## Visualize first two principal components

::: columns
::: {.column}

```{r}
#| label: pca-plot
#| eval: false
starbucks <- starbucks |> 
  mutate(pc1 = starbucks_pc_matrix[,1], 
         pc2 = starbucks_pc_matrix[,2])
starbucks |> 
  ggplot(aes(x = pc1, y = pc2)) +
  geom_point(alpha = 0.5) +
  labs(x = "PC 1", y = "PC 2")
```

* Principal components are not interpretable

* Make a **biplot** with arrows showing the linear relationship between one variable and other variables

:::

::: {.column}

```{r}
#| ref-label: pca-plot
#| echo: false
#| fig-height: 9
```

:::
:::


## Making PCs interpretable with biplots

* Biplot displays both the space of observations and the space of variables

::: columns
::: {.column}

```{r}
#| label: pca-biplot
#| eval: false
library(factoextra)
# fviz_pca_var(): projection of variables
# fviz_pca_ind(): display observations with first two PCs
starbucks_pca |> 
  fviz_pca_biplot(label = "var",
                  alpha.ind = 0.25,
                  alpha.var = 0.75,
                  labelsize = 5,
                  col.var = "darkblue",
                  repel = TRUE)
```

* Interpret a biplot

  * Arrow direction

  * Arrow angles
  
  * Arrow length

:::


::: {.column}

```{r}
#| ref-label: pca-biplot
#| echo: false
#| fig-height: 10
```

:::
:::

## Interpreting a biplot

::: columns
::: {.column}

Interpreting the **direction of a particular arrow**

* For example, far left arrow for `caffeine_mg`: as `caffeine_mg` increases, $Z_1$ and $Z_2$ tend to decrease

  * i.e., within the definition of $Z_1$ and $Z_2$, the coefficient for `caffeine_mg` is negative

* In contrast, arrow for `serv_size_m_l` points to upper right: as `serv_size_m_l` increases,<br>both $Z_1$ and $Z_2$ also increase

:::


::: {.column}

```{r}
#| ref-label: pca-biplot
#| echo: false
#| fig-height: 10
```


:::
:::

## Interpreting a biplot

::: columns
::: {.column}

**The angle of between the vectors is indicative of the correlation between different variables**

* $90^{\circ}$ (right angle): uncorrelated<br>(e.g. `serv_size_m_l` & `saturated_fat_g`)

* $< 90^{\circ}$ (vectors in similar directions):<br>positively correlated<br>(e.g. `sugar_g` & `total_carbs_g`)

* $> 90^{\circ}$ (vectors in different directions):<br>negatively correlated<br>(e.g. `caffeine_mg` & `calories`)

:::


::: {.column}

```{r}
#| ref-label: pca-biplot
#| echo: false
#| fig-height: 10
```
:::
:::

## Interpreting a biplot

::: columns
::: {.column}

**The arrow length indicates how strongly related the principal components are with the individual variables**

* For example, `serv_size_m_l` has a fairly long arrow because it has a large positive coefficient in the rotation matrix (see below)

* Meanwhile, `caffeine_mg` has a relatively short arrow because its coefficients are relatively small.

:::


::: {.column}

```{r}
#| ref-label: pca-biplot
#| echo: false
#| fig-height: 10
```

:::
:::

## Intuition on how many PCs to use

```{r}
summary(starbucks_pca)
```

. . .

* PC1 accounts for the most variation in our data, PC2 accounts for the next most, and so on

. . .

* Each time we add a new PC dimension, we capture a higher proportion of the variability/information in the data

. . .

* But that increase in proportion decreases for each new dimension we add (i.e., additional PCs will add smaller and smaller variance)

. . .

* **It is recommended to keep adding PCs until the marginal gain levels off (i.e. decreases to the point that it isn’t too beneficial to add another dimension to the data)**


## Create a scree plot (or elbow plot)

::: columns
::: {.column}

```{r}
#| eval: false
starbucks_pca |> 
  fviz_eig(addlabels = TRUE) +
  geom_hline(yintercept = 100 * (1 / ncol(starbucks_pca$x)), 
             linetype = "dashed", 
             color = "darkred")
```

```{r}
#| echo: false
#| fig-height: 7.5
starbucks_pca |> 
  fviz_eig(addlabels = TRUE) +
  geom_hline(yintercept = 100 * (1 / ncol(starbucks_pca$x)), 
             linetype = "dashed", 
             color = "darkred") +
  theme(text = element_text(size = 20))
```


:::

::: {.column}

* Shows proportion of variation explained by each PC

* Visual rule-of-thumb: look for the "elbow" (where the proportion of variation starts to become flat)

* A heuristic rule of thumb: horizontal line at $1/p$

<br>
<br>

* The scree plot suggests keeping the first 3 PCs<br>(could even go up to first 5 PCs for another substantial drop in the elbow)

:::
:::

## Discussion

::: columns
::: {.column}

```{r}
#| eval: false
starbucks_pca |> 
  fviz_eig(addlabels = TRUE) +
  geom_hline(yintercept = 100 * (1 / ncol(starbucks_pca$x)), 
             linetype = "dashed", 
             color = "darkred")
```

```{r}
#| echo: false
#| fig-height: 7.5
starbucks_pca |> 
  fviz_eig(addlabels = TRUE) +
  geom_hline(yintercept = 100 * (1 / ncol(starbucks_pca$x)), 
             linetype = "dashed", 
             color = "darkred") +
  theme(text = element_text(size = 20))
```

:::

::: {.column}

* Suppose we decide to use first 3 PCs (to capture satisfactory amount of the variation in the data)

* A standard biplot displays only the first 2 PCs, meaning that some information is omitted

  * In this case, the omitted information corresponds to roughly 30% of the total variation (i.e., the variation captured by PCs 3--11)
  
  * About 1/3 of this remaining information is captured by PC3 (not plotted)
  
* In principle, we should plot 3 quantitative variables (e.g., point size, transparency, 3D scatterplot, etc.)

* Alternatively, we could make three scatterplots (one for each pair of PCs)
:::
:::


## Remember the spelling...

<center>

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">folks. it&#39;s _principal_ (not principle) components analysis.</p>&mdash; Stephanie Hicks, PhD (@stephaniehicks) <a href="https://x.com/stephaniehicks/status/1626428875323367424?ref_src=twsrc%5Etfw">February 17, 2023</a></blockquote> <script async src="https://platform.x.com/widgets.js" charset="utf-8"></script>

</center>

## Appendix: PCA output

```{r}
# str(starbucks_pca)
```

Examine the output after running `prcomp()`

* `starbucks_pca$sdev`: singular values ($\sqrt{\lambda_j}$)

* `starbucks_pca$rotation`: loading matrix ($V$)

*  `starbucks_pca$x`: principal component scores matrix ($Z=XV$)

Can use the `broom` package for tidying `prcomp()`

```{r}
#| eval: false
tidy(starbucks_pca, matrix = "eigenvalues") # equivalent to starbucks_pca$sdev
tidy(starbucks_pca, matrix = "rotation") # equivalent to starbucks_pca$rotation
tidy(starbucks_pca, matrix = "scores") # equivalent to starbucks_pca$x
```

## Appendix: Proportion of variance explained

```{r}
library(broom)
starbucks_pca |>
  tidy(matrix = "eigenvalues") |>
  ggplot(aes(x = PC, y = percent)) +
  geom_line() + 
  geom_point() +
  geom_hline(yintercept = 1 / ncol(starbucks_feat), color = "darkred", linetype = "dashed") +
  scale_x_continuous(breaks = 1:ncol(starbucks_pca$x))
```

## Appendix: Cumulative proportion of variance explained

```{r}
library(broom)
starbucks_pca |>
  tidy(matrix = "eigenvalues") |>
  ggplot(aes(x = PC, y = cumulative)) +
  geom_line() + 
  geom_point() +
  scale_x_continuous(breaks = 1:ncol(starbucks_pca$x))
```

## Appendix: Singular value decomposition (SVD)

$$
X = U D V^T
$$

* $U$ and $V$: matrices containing the left and right **singular vectors of scaled matrix $X$**

* $D$: diagonal matrix of the **singular values**

. . .

* SVD simplifies matrix-vector multiplication as **rotate, scale, and rotate again**

. . .

$V$: **loading matrix** for $X$ with $\phi_{j}$ as columns

* $Z = X  V$: PC matrix

. . .

Bonus: **Eigenvalue decomposition** (or spectral decomposition)

* $V$: **eigenvectors** of $X^TX$ (covariance matrix)

* $U$: **eigenvectors** of $XX^T$

* The singular values (diagonal of $D$) are square roots of the **eigenvalues** of $X^TX$ or $XX^T$

* Meaning that $Z = UD$

## Appendix: Eigenvalues guide dimension reduction

We want to choose $p^* < p$ such that we are explaining variation in the data

. . .

Eigenvalues $\lambda_j$ for $j \in 1, \dots, p$ indicate **the variance explained by each component**

* $\displaystyle \sum_j^p \lambda_j = p$, meaning $\lambda_j \geq 1$ indicates $\text{PC}j$ contains at least one variable's worth in variability
  
* $\displaystyle \frac{\lambda_j}{p}$: proportion of variance explained by $\text{PC}j$
  
* Arranged in descending order so that $\lambda_1$ is largest eigenvalue and corresponds to PC1
  
. . .
  
* Compute the cumulative proportion of variance explained (CVE) with $p^*$ components $$\text{CVE}_{p^*} = \sum_j^{p*} \frac{\lambda_j}{p}$$

Use a **scree plot** to plot eigenvalues and guide choice for $p^* <p$ by looking for "elbow" (rapid to slow change)