---
title: "Visualizing text data"
subtitle: "<br>36-315: Statistical Graphics and Visualization, Summer 2026"
footer:  "[36-315-summer26.netlify.app](https://36-315-summer26.netlify.app)"
format:
  revealjs:
    embed-resources: true
    height: 900
    width: 1600
    theme: theme.scss
    smaller: true
    slide-number: c/t
    linestretch: 1.4
    code-line-numbers: false
    html-math-method:
      method: mathjax
      url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
set.seed(8)
```

## Text data example

* The Office, Season 4 Episode 13: [Dinner Party](https://en.wikipedia.org/wiki/Dinner_Party_(The_Office))

* Access via the [`schrute` package](https://cran.r-project.org/web/packages/schrute/vignettes/theoffice.html)

```{r}
library(tidyverse)
theme_set(theme_light())
```

```{r}
#| echo: false
theme_set(theme_light(base_size = 18))
```

```{r}
library(schrute)
transcripts <- theoffice |>
  filter(season == 4, episode == 13) |>
  select(index, character, text)
head(transcripts)
```

## Bag of words representation of text

Most common way to store text data: **document-term matrix** (DTM)

|              | Word 1   | Word 2   | $\dots$  | Word $J$ |
| ------------ | -------- | -------- | -------- | -------- |
| Document $1$ | $w_{11}$ | $w_{12}$ | $\dots$  | $w_{1J}$ |
| Document $2$ | $w_{21}$ | $w_{22}$ | $\dots$  | $w_{2J}$ |
| $\dots$      | $\dots$  | $\dots$  | $\dots$  | $\dots$  |
| Document $N$ | $w_{N1}$ | $w_{N2}$ | $\dots$  | $w_{NJ}$ |

* $w_{ij}$: count of word $j$ in document $i$, or term frequency

. . .

To reduce number of columns:

* **Stop words**: remove extremely common, functional words (e.g., the, and, is, in, of, etc.)

* **Stemming**: reduce all words to their "stem"

  * Example: compute, computer, computed, computing, computation $\rightarrow$ comput


## Tokenize text into long format

* Convert raw text into long, tidy table with one-token-per-document-per-row

  * A **token**: a unit of text (typically a word)
  
* Easy to convert text into DTM format using [`tidytext` package](https://cran.r-project.org/web/packages/tidytext/vignettes/tidytext.html)  
  
```{r}
library(tidytext)
transcripts_tokens <- transcripts |>
  unnest_tokens(word, text)
head(transcripts_tokens)
```

## Remove stop words

::: columns
::: {.column}

* `tidytext` provides `stop_words` data

```{r}
head(stop_words)
```

:::
::: {.column}

* Remove stop words from all tokens

```{r}
transcripts_tokens <- transcripts_tokens |>
  filter(!(word %in% stop_words$word))
head(transcripts_tokens)
```

:::
:::

## Visualize distribution of most common words

```{r}
#| code-fold: true
transcripts_tokens |> 
  count(word, sort = TRUE) |> 
  filter(n > 5) |> 
  mutate(word = fct_reorder(word, n)) |> 
  ggplot(aes(n, word)) +
  geom_col() +
  labs(y = NULL)
```

## Stemming

* Use [`SnowballC` package](https://cran.r-project.org/web/packages/SnowballC/SnowballC.pdf) and `wordStem()` function to perform stemming

```{r}
library(SnowballC)
transcripts_tokens <- transcripts_tokens |>
  mutate(stem = wordStem(word))
head(transcripts_tokens)
```

## Tokenize by n-gram

* An **n-gram** is a sequence of n adjacent items (e.g., words, letters, symbols, etc.)

. . .

::: columns
::: {.column}

* Consider pairs of two consecutive words ("bigrams") instead of single words

```{r}
transcripts_bigrams <- transcripts |>
  unnest_tokens(bigram,text, token = "ngrams", n = 2) |> 
  filter(!is.na(bigram))
head(transcripts_bigrams)
```

:::
::: {.column}

* Remove stop words: `separate()` into two words, then `filter()`, then `unite()`

```{r}
transcripts_bigrams <- transcripts_bigrams |> 
  separate(bigram, c("word1", "word2"), sep = " ") |> 
  filter(!word1 %in% stop_words$word,
         !word2 %in% stop_words$word) |> 
  unite(bigram, word1, word2, sep = " ")
head(transcripts_bigrams)
```

:::
:::

. . .

* Can perform the same frequency analysis as for single words

## Word clouds

* A **word cloud** displays all words mentioned across documents, where more common words are larger

  * First, compute the word counts $$w_{\cdot 1} = \sum_{i=1}^N w_{i1} \hspace{0.1in} \dots \hspace{0.1in} w_{\cdot J} = \sum_{i=1}^N w_{iJ}$$

  * Then, the size of word $j$ is proportional to $w_{\cdot j}$

. . .

* Create word clouds in `R` using [`wordcloud` package](https://cran.r-project.org/web/packages/wordcloud/wordcloud.pdf)

* `wordcloud()` function takes in two main inputs: `words`(vector of unique words) and `freq` (vector of frequencies)

## Word clouds

```{r}
#| code-fold: true
token_summary <- transcripts_tokens |>
  count(stem)
library(wordcloud)
wordcloud(words = token_summary$stem, 
          freq = token_summary$n, 
          random.order = FALSE, # place biggest words in center
          max.words = 100, 
          colors = brewer.pal(8, "Dark2"))
```

## Comparison clouds

* Goal: create word clouds to compare two different collections of documents $\mathcal{A}$ and $\mathcal{B}$

. . .

* Construct vectors of total words for each collection

  * $\mathbf{w}^{\mathcal{A}} = (w_{\cdot 1}^{\mathcal{A}}, \dots, w_{\cdot J}^{\mathcal{A}})$

  * $\mathbf{w}^{\mathcal{B}} = (w_{\cdot 1}^{\mathcal{B}}, \dots, w_{\cdot J}^{\mathcal{B}})$

. . .

* Consider the $j$th word (say "dinner")

  * If $w_{\cdot j}^{\mathcal{A}}$ is large, then "dinner" is large in the word cloud for $\mathcal{A}$

  * If $w_{\cdot j}^{\mathcal{B}}$ is large, then "dinner" is large in the word cloud for $\mathcal{B}$

  * But if both are large, this does not indicate whether $w_{\cdot j}^{\mathcal{A}}$ or $w_{\cdot j}^{\mathcal{B}}$ is bigger

## Comparison clouds

* For word $j$, compute $\bar{w}_{\cdot j} = \textsf{average}(w_{\cdot j}^{\mathcal{A}}, w_{\cdot j}^{\mathcal{B}})$

. . .

* Compute $w_{\cdot j}^{\mathcal{A}} - \bar{w}_{\cdot j}$ and $w_{\cdot j}^{\mathcal{B}} - \bar{w}_{\cdot j}$

. . .

* If $w_{\cdot j}^{\mathcal{A}} - \bar{w}_{\cdot j}$ is very positive, make it large for the $\mathcal{A}$ word cloud<br>If $w_{\cdot j}^{\mathcal{B}} - \bar{w}_{\cdot j}$ is very positive, make it large for the $\mathcal{B}$ word cloud

## Comparison clouds

```{r}
#| code-fold: true
#| fig-height: 4
#| fig-width: 10
michael_jan_summary <- transcripts_tokens |>
  count(character, stem) |> 
  filter(character %in% c("Michael", "Jan"))

library(reshape2)
michael_jan_summary |>
  acast(stem ~ character, value.var = "n", fill = 0) |>
  comparison.cloud(colors = c("red", "blue"), max.words = 100, scale = c(2, 0.1), title.size = 2)
```

## Term frequency-inverse document frequency (tf-idf)

* Goal: compare text across documents (e.g., The Office characters)

* Determine the most unique words for each document using **tf-idf**

* Approach: downweight words that occur frequently across all documents

. . .

* **Inverse document frequency (idf)**: For word $j$, compute

$$\textsf{idf}_j = \log \frac{N}{N_j},$$ where $N$ is number of documents and $N_j$ is number of documents containing word $j$

* **Term frequency-inverse document frequency (tf-idf)**

$$\textsf{tf-idf}_j = \textsf{tf}_j \times \textsf{idf}_j = w_{ij} \times \textsf{idf}_j$$

. . .

* The higher the tf-idf value (unitless), the more unique the word is in the document

## tf-idf example

Use `bind_tf_idf()` from `tidytext` package

```{r}
transcripts_tfidf <- transcripts_tokens |>
  count(character, stem) |> 
  bind_tf_idf(stem, character, n) 
head(transcripts_tfidf)
```

## Visualize most unique words according to tf-idf

```{r}
#| code-fold: true
#| fig-height: 8
transcripts_tfidf |>
  filter(character %in% c("Michael", "Jan", "Jim", "Pam")) |>
  group_by(character) |>
  slice_max(tf_idf, n = 7, with_ties = FALSE) |>
  mutate(stem = reorder_within(stem, tf_idf, character)) |>
  ggplot(aes(x = tf_idf, y = stem)) +
  geom_col() +
  scale_y_reordered() +
  facet_wrap(~ character, ncol = 2, scales = "free") +
  labs(x = "tf-idf", y = NULL)
```

## Sentiment analysis

* Goal: understand the emotional intent or sentiment of words

. . .

* Lexicon-based approach

  * Find a sentiment lexicon/dictionary
  
  * Compute summary (e.g., average sentiment)
  
. . .
  
* Different types of sentiment lexicons
  
  * Positive or negative
  
  * Numerical score (e.g., from −5 to +5) 
  
  * Specific emotions (e.g., sadness, anger, joy, etc.)
  
## Sentiment analysis example

::: columns
::: {.column}

* Consider the [Bing sentiment lexicon](https://www.cs.uic.edu/~liub/FBS/sentiment-analysis.html)<br>(positive/negative)

```{r}
bing_sentiment <- get_sentiments("bing")
head(bing_sentiment)
```

:::
::: {.column}

* Join sentiment to token table

```{r}
transcripts_sentiment <- transcripts_tokens |> 
  inner_join(bing_sentiment)
head(transcripts_sentiment)
```

:::
:::

* Make different graphics to visualize sentiment

## Sentiment word clouds

```{r}
#| code-fold: true
#| fig-height: 4
#| fig-width: 9
transcripts_sentiment |> 
  count(word, sentiment) |> 
  acast(word ~ sentiment, value.var = "n", fill = 0) |> 
  comparison.cloud(colors = c("midnightblue", "red"), max.words = 100, scale = c(2, 0.5), title.size = 1)
```
