---
title: "YOUR PROJECT TITLE HERE"
subtitle: "36-315: Statistical Graphics and Visualization, Summer 2026"
author: "YOUR NAME HERE"
format:
  html:
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

---

# A quick tutorial

**(Feel free to remove this section when you submit)**

This a Quarto document. 
To learn more about Quarto see <https://quarto.org>.
You can use the Render button to see what it looks like in HTML.

### Text formatting

Text can be bolded with **double asterisks** and italicized with *single asterisks*. 
Monospace text, such as for short code snippets, uses `backticks`.
(Note these are different from quotation marks or apostrophes.) Links are
written [like this](http://example.com/).

Bulleted lists can be written with asterisks:

* Each item starts on a new line with an asterisk.
* Items should start on the beginning of the line.
* Leave blank lines after the end of the list so the list does not continue.

### Code blocks

Code blocks are evaluated sequentially when you hit Render. 
As the code runs, `R` prints out which block is running, so naming blocks is useful if you want to know which one takes a long time. 
After the block name, you can specify chunk options.
For example, `echo` controls whether the code is printed in the document. 
By default, output is printed in the document in monospace:

```{r}
#| echo: false
head(mtcars)
```

### Figures

If a code block produces a plot or figure, this figure will automatically be inserted inline in the report. That is, it will be inserted exactly where the code block is.

```{r}
#| fig-width: 5
#| fig-height: 3.5
#| fig-cap: "This is a caption. It should explain what's in the figure and what's interesting about it. For instance: There is a negative, strong linear correlation between miles per gallon and horsepower for US cars in the 1970s."

library(ggplot2)
mtcars |> 
  ggplot(aes(x = mpg, y = hp)) +
  geom_point() +
  labs(x = "Miles per gallon",
       y = "Horsepower")
```

Notice the use of `fig-width` and `fig-height` to control the figure's size (in inches). 
These control the sizes given to `R` when it generates the plot, so `R` proportionally adjusts the font sizes to be large enough.
